Evolution of Processes in Delivering Reliable Software in the European Scientific Arena

### 2018-06-13 paper-ieee-trans.tex
started adding background section

Major developments carried throughout the years where focused on providing new
functionalities, while in the early days, the adoption of software engineering methodologies had a lower priority.
The released software was less tested, resulting in a higher rate of
rollbacks and patches applied to production systems, compared to the later projects. Thereby, the stability and
% Comment
%P1 C2 L18: Claiming that stability and reliability was regularly compromised is quite a bold statement -- it needs at least some citations, and preferably some explanation of what forms this compromise took for a few prominent cases.
reliability of the e--Infrastructures was often compromised, affecting
negatively their exploitation by the scientific communities \cite{aiftimiei}.
% Comments
%P1 C2 L24: Was this realisation imposed "from above" e.g. by coordinated EC/ERC oversight, or was it "emergent" from the demands of the independent scientific communities?
It became then apparent the need for a better balance between the addition of new features
and the reliability of the released software. Consequently, incoming projects progressively
devoted a substantial part of the software lifecycle to the improvement of the quality of the
software being delivered. At this point, evolving software engineering practices were considered
and gradually adopted to define and implement quality procedures, the organization of
cross--functional teams and deployment of pilot testbeds.

@@

The reminder of this paper is
organized as follows. Section~\ref{sec:ev} details the evolution of research on
software reliability in EC--projects. Section~\ref{sec:ntsr} provides
information about new research trends in software reliability based upon the
experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section~\ref{sec:con}.

@@

\subsection{Software quality validation and delivery automation}
\label{sec:devops}
% Comment
%P5 C2 L21: Define DevOps!
Tackling the granularity of the former SQA requirements -- most of which accounted in a per--change basis --
is hardly achieved without the aid of automation. The introduction of automation to validate the quality
requirements increases the overall reliability of the produced software: automated testing is more 
time-efficient, leading to higher coverages and increased defect detection, when compared with a manual
approach \cite{rafi2012}.

As any new change, involving a new feature or fix, passes through the automated
SQA machinery, the software is always ready to be released. As a result, the often compromised harmonization
between the continouous release of new features -- pushed by the development teams -- and the maintenance of
stable production systems -- demanded by users and resource providers -- can be guaranteed. The DevOps culture
theorizes and provides practical solutions that aim at unifying both software development (dev) and software
operations (ops) teams, by emphasizing the SQA techniques to avoid infrastructure disruption whenever new
developments are deployed into production systems. The following DevOps approaches have been progressively
adopted throughout the project's lifetime, being one of its major outcomes in terms of software reliability.

@@

\subsubsection{Continuous Delivery}
% Comment from reviewer
%Please add a reference.
As DevOps suggests, frequent releases positively affect the reliability of the software as
they reduce the time in which a given defect is exposed or has an impact, allowing the 
development teams to act promptly based on the regular user feedback \cite{chen2015}. The software updates of
{\sl INDIGO--DataCloud} products, that has been taking place since the second major release, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI--validated
software components are automatically delivered in online repositories without the need of
any human interaction but ultimately requiring expert supervision to validate the results.

@@

>>>>

Major developments carried throughout the years were focused on providing new
functionalities, whereas in the early days the adoption of software engineering methodologies had a lower priority.
The released software was less tested, resulting in a higher rate of
rollbacks and patches applied to production systems, compared to the later projects. Thereby, the stability and
% Comment
%P1 C2 L18: Claiming that stability and reliability was regularly compromised is quite a bold statement -- it needs at least some citations, and preferably some explanation of what forms this compromise took for a few prominent cases.
reliability of the e--Infrastructures was often compromised, affecting
negatively their exploitation by the scientific communities \cite{aiftimiei}.
% Comments
%P1 C2 L24: Was this realisation imposed "from above" e.g. by coordinated EC/ERC oversight, or was it "emergent" from the demands of the independent scientific communities?
It became then apparent the need for a better balance between the addition of new features
and the reliability of the released software. Consequently, incoming projects progressively
devoted a substantial part of the software lifecycle to the improvement of the quality of the
software being delivered. At this point, evolving software engineering practices were considered
and gradually adopted to define and implement quality procedures, the organization of
cross--functional teams and deployment of pilot testbeds.

@@

The reminder of this paper is
organized as follows. Section~\ref{sec:b} summaries the essential context of the research problem. Section~\ref{sec:ev} details the evolution of research on
software reliability in EC--projects. Section~\ref{sec:ntsr} provides
information about new research trends in software reliability based upon the
experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section~\ref{sec:con}.

\section{Background}
\label{sec:b}

In this section, we detail the level of knowledge in delivering reliable software before and after the EC-founded projects.
We divide literature into four different domains that intersect this work. Firstly, we describe the software development process showing how the various phases reflect on software quality attributes. Next, we describe the impact of software requirements on software quality. In the last, we introduce metrics and measurements. In the last, we talk about software testing and code coverage.

\emph{Metrics and measurements.} Metrics and measurements have an important role in the software development process \cite{fenton, capers}.
Metrics are used to improve the ability to control, identify and measure the essential parameters of software  during  its  development.
leading to improve the results of the software product \cite{goodman}.
With a correct usage of  software  metrics,  it  is possible to determine which parts of software require to be changed or modified. In literature different types of metrics exist able to provide information on various software aspects, such as complexity \cite{mccabe} and object orientation \cite{chidamber, lorenz}. Often they have similar definitions, as a consequence they can be implemented in different ways according to the interpretation of their definition. This is the reason why it has been always a challenge to select a proper software metrics tool.

\emph{Testing.} Testing is an important factor for software quality and reliability improvement. In literature there are many types of software testing \cite{myers} that aim at finding errors, bugs and faults during the software development process, possibly in the early stages. Test criteria focus on measuring the coverage of the test suite

@@

\subsection{Software quality validation and delivery automation}
\label{sec:devops}
% Comment
%P5 C2 L21: Define DevOps!
Tackling the granularity of the former SQA requirements -- most of which accounted in a per--change basis --
is hardly achieved without the aid of automation. The introduction of automation to validate the quality
requirements increases the overall reliability of the produced software: automated testing is more 
time-efficient, leading to higher code coverages and increased defect detection, when compared with a manual
approach \cite{rafi2012}.

As any new change, involving a new feature or fix, passes through the automated
SQA machinery, the software is always ready to be released. As a result, the often compromised harmonization
between the continouous release of new features -- pushed by the development teams -- and the maintenance of
stable production systems -- demanded by users and resource providers -- can be guaranteed. The DevOps culture
theorizes and provides practical solutions that aim at unifying both software development (dev) and software
operations (ops) teams, by emphasizing the SQA techniques to avoid infrastructure disruption whenever new
developments are deployed into production systems. 
The incremental adoption of DevOps approaches throughout the project's lifetime was a key player towards
software reliability. Starting with a CI scenario at the first stages of the project, the software delivered
was eventually validated and distributed using CD pipelines.

@@

\subsubsection{Continuous Delivery}
% Comment from reviewer
%Please add a reference.
As DevOps suggests, frequent releases positively affect the reliability of the software as
they reduce the time in which a given defect is exposed or has an impact, allowing the 
development teams to act promptly based on the regular feedback \cite{chen2015}. The software updates of
{\sl INDIGO--DataCloud} products, that have been taking place since the second major release, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI--validated
software components are automatically delivered in online repositories without the need of
any human interaction but ultimately requiring expert supervision to validate the results.

@@

\bibitem{fenton}
N.  Fenton,  and  S.  L.  Pfleeger.
Software  Metrics:  A  Rigorous  and  Practical Approach,
\emph{International Thomson Computer
Press}, London, UK, second edition, 1997.

\bibitem{capers}
J. T. Capers.
Applied Software Measurement
\emph{McGrawHill}, 1991.

\bibitem{goodman}
P. Goodman.
Practical Implementation of Software Metrics, 
\emph{McGraw Hill}, London, 1993.

\bibitem{mccabe}
T. J. McCabe.
A   Complexity   Measure,
\emph{IEEE   Transactions   on   Software Engineering},
Vol. 2, No. 4, pp. 308-320, 1976.

\bibitem{chidamber}
S.  R.  Chidamber  and  C.  E  Kemerer.
A  metrics  suite  for  Object--oriented  design,
\emph{IEEE Trans. Software Engineering}, 
June 1994.

\bibitem{lorenz}
M. Lorenz and J. Kidd.
Object-Oriented Software Metrics: A Practical Approach,
\emph{IEEE publications}.

\bibitem{myers}
G. J. Myers and C. Sandler and T. Badgett.
The art of software testing.
\emph{John Wiley \& Sons}, 2011.
